Type Evolution in Programming Systems

### 2024-05-24 paper.tex
Rewrite framework, migrate some of Elm

\usepackage{csquotes}
\usepackage{changepage}
\usepackage{multicol}
\usepackage{ccicons}
\usepackage[many]{tcolorbox}
\lstdefinelanguage[programming]{TeX}[AlLaTeX]{TeX}{%
  deletetexcs={title,author,bibliography},%
  deletekeywords={tabular},
  morekeywords={abstract},%
  moretexcs={chapter},%
  moretexcs=[2]{title,author,subtitle,keywords,maketitle,titlerunning,authorinfo,affiliation,authorrunning,paperdetails,acks,email},
  moretexcs=[3]{addbibresource,printbibliography,bibliography},%
}%
\lstset{%
  language={[programming]TeX},%
  keywordstyle=\firamedium,
  stringstyle=\color{RosyBrown},%
  texcsstyle=*{\color{Purple}\mdseries},%
  texcsstyle=*[2]{\color{Blue1}},%
  texcsstyle=*[3]{\color{ForestGreen}},%
  commentstyle={\color{FireBrick}},%
  escapechar=`,}
\newcommand*{\CTAN}[1]{\href{http://ctan.org/tex-archive/#1}{\nolinkurl{CTAN:#1}}}
%%

\newcounter{challengen}
\newcommand{\challenge}[1]{\subsection*{\addtocounter{challengen}{1} Challenge \#\thechallengen: #1}}

\DeclareRobustCommand{\frameworkbox}[2][gray!15]{
\begin{tcolorbox}[breakable,left=3pt,right=3pt,top=3pt,bottom=3pt,colback=#1,colframe=#1,parbox=false,
  width=\dimexpr\textwidth\relax,enlarge left by=0mm,boxsep=5pt,arc=0pt,enlarge top by=0.5em,%enlarge bottom by=0.0em,
  outer arc=0pt]\setlength{\parskip}{0.5em}\setlength{\parindent}{0em}\textbf{\sffamily Framework perspective.}\quad #2
\end{tcolorbox}}

@@

Many past improvements in the practice of programming worked by speeding up feedback loops,
both in collaboration with other programmers (as in collaborative development tools
\cite{goldman2011real,kurniawan2015coder,replit}) and in individual's interaction with the
programming environment (as in live programming \cite{rein2018exploratory}).

To further tighten the feedback loop and make programming more effective, future programming
systems will need to enable programmers to collaborate at a fine-grained level, letting them
adopt individual changes created by their collaborators (as in local-first software
\cite{localfirst}). To provide immediate feedback, future programming systems will need to
transcend treating state as ephemeral, tied to the lifetime of a running program and reset when the
program is edited.

We need to shift our focus to \emph{programming systems} \cite{techdims} where program state
is persistent and co-evolves with the behavior and data of the system. This is already the case
in collaborative applications with state persistent in a database, image-based systems like
Smalltalk and live programming environments with hot-reloading where the program can be edited
without resetting all state.

While we are firmly convinced that more collaborative and live programming systems are the
future of programming, there are a number of problems that need to be resolved in order for
such systems to become a reality. The key problem is of \emph{type evolution}, i.e., how to
handle the change of schema of the data program works with in a way that preserves integrity
constraints, existing data and the logic of code.

Type evolution is not a new problem. It exists in various forms in a range of existing systems.
In the database world, the problem of schema evolution is well-studied \cite{erhard06}.
It typically focuses on ensuring data integrity, although co-evolution of database structure
and code that manipulates the database has also been studied \cite{Cleve2006,wang2019synthesizing}.
In live programming, the problem is known as hot code reloading \cite{barenz2020essence,beckmann2021shortening},
although it typically focuses on updating transient state and ignores the problem of more complex
data migrations. The problem is equally studied in the bidirectional transformation (BX)
communities \cite{czarnecki2009bidirectional} and in work on model-driven development
\cite{metamodels,alanen2003difference}.

Type evolution is an important challenge for future programming systems. But looking at the problem
from the perspective of programming systems can also offer a new perspective that
unfies the many different aspects of the problem partially explored in other areas. A programming
system may have permanent data (as in a database) whose structure can change, requiring
corresponding change to data and code. A programming system may have a transient state conforming
to a type structure that may change during live programming. A programming system may also need
to allow programmers to independently work on different versions of a program.

In this paper, we draw connections between these different kinds of type evolution, pointing out
essential shared structure. By doing so, we aim to motivate further research on this important
problem and encouraging knowledge sharing across subfields. Our contributions are twofold.
First, we present a shared conceptual framework for thinking about the relationship between
schema, code, and data in a programming system. Second, we present five case studies that use
the framework to study type evolution in a highly diverse set of real-world programming scenarios.

\begin{figure}[t]
\centering
\vspace{-1em}
\includegraphics[width=0.45\textwidth]{figures/layers.png}
\caption{Pace layers from Stewart Brand's classic \emph{How Buildings Learn} \cite{Brand95}.
  Different layers of a building change at different pace. The geographic
  site of a building is eternal, its structure remains stable for decades, space plan
  changes every couple of years.}
\label{fig:layers}
\end{figure}

\section{Type Evolution}
To paraphrase Stewart Brand \cite{Brand95}, ``almost no programs evolve well. But all programs
(except for monuments) evolve anyway, however poorly, because the usages in and around them are
changing constantly.'' Programs evolve both while they are being created and during their operation
in response to new needs. As with buildings (Figure~\ref{fig:layers}), different aspects of
program evolve at different pace. And ``because of the different rates of change of its components,
a program is always tearing itself apart.''

\subsection{Types, Code and Data}

As is the case for buildings, making a change at some layers (skin or stuff) is relatively easy,
but a change at a more basic layer (structure) has cascading effects on layers that depend on it.
In the case of programs, a more basic layer is a structure that defines the shape of data and code.
We refer to this structure as \emph{types}, although we do not use the term in an precise technical
sense. In our conceptual framework, types may be data types that define the shape of data. They
may be explicitly declared in code and checked statically, as well as dynamically checked or
even implicit. The key property of types for us is that they define the shape of other layers
of a program, such as data and code. As a result, a change of program types requires a
corresponding change in program data and code.

@@

The second dimension of challenges that we consider is non-local. A program may have multiple
variants that exist independently and need to be synchronized. The difficulty of the problem
depends on what kind of synchronization we want to support. If all variants of the program should
eventually adopt all changes (converge), the complexity is lesser. If we allow user to apply
only certain changes from other users (divergence), the complexity is greater. A typical case
may be that the types and schema of a program should converge, but data may diverge.

\begin{figure}[t]
\centering
\includegraphics[width=0.65\textwidth]{figures/arrows.png}
\caption{When a type changes from $T_1$ to $T_2$, a corresponding change needs to take place
  in data and code whose shape is defined by type. The challenge is finding a suitable
  corresponding transformation (semi-)automatically.}
\label{fig:change}
\end{figure}

@@

@@

\textcolor{red}{TODO: got as far as here so far...}

@@

\paragraph{Dimensions.}
The framework for talking about schema change used in this paper thus distinguishes between
two dimensions:

\begin{itemize}
  \item The \emph{local dimension} is concerned with difficulties that a change causes
    across different program pace layers that exist within a single program.
  \item The \emph{non-local dimension} is concerned with the difficulties that a change causes
    across different program variants that exist independently.
\end{itemize}

\paragraph{Program pace layers.} As with buildings (Figure~\ref{fig:layers}), programs consist
of multiple pace layers that change at a different rate. As with buildings, making a change
to a more permanent layer also typically affects all the less permanent layers. Making a change
at a more permanent layer is thus more challenging. Alas, changing more permanent structures in
programming is frequently necessary. In this paper, we distinguish between four layers.

\begin{itemize}
\item \emph{State} -- transient state of a program. This is typically never shared across
  multiple variants of a program and it can often be discarded when a change occurs.
\item \emph{Data} -- data collected by the program. This represents a more permanent structure
  that is typically stored in a database and outlives program restarts.
\item \emph{Code} -- program logic. This layer changes whenever any aspect of the program logic
  is modified. A change typically affects transient state, but may not affect data.
\item \emph{Schema} -- the structures of program data. A change at this layer typically
  requires adapting code to work with the new structure and also updating data and state.
\end{itemize}

\paragraph{Program variants.} We are interested in programs that are distributed across multiple
environments as, for example, in local-first software. As discussed earlier, one aspect of this
dimension is whether all variants of a program eventually converge or whether it is possible to
maintain a diverging version. It is also useful to distinguish the directions in which a
change may flow, i.e., can a change originate in any variant or is there a single centralized
source of change. Note that each of this characteristics may  be associated with individual pace
layers of a program.

\begin{itemize}
\item \emph{Convergence} vs. \emph{Divergence}. In the convergence model, all program variants
  eventually adopt all changes. It may not be possible to adopt a particular change before adopting
  other changes it depends on. In the divergence model, a user may choose only particular changes
  they want to adopt.
\item \emph{Centralized} vs. \emph{Decentralized}. In the centralized model, changes (at a
  specific layer) can only originate from a particular source. In the decentralized model,
  changes can be done on any of the multiple co-existing variants of a program.
\end{itemize}

\paragraph{Challenge problems.}
The framework outlined above offers a wide range of configurations and one could conceive of
a challenge problem for many of those configurations. We can start with a change at each of the
four layers, consider each of the four configurations with respect to program variants (and also
consider different behaviour at all the affected less permanent layers). The aim of this paper
is, however, not to enumerate all options. Instead, our goal is to present a number of typical
and often encountered challenge problems. We use the above framework to more precisely define
the nature of the challenge problem. We also often use the above framework to discuss variations
on the particular challenge.

@@

@@

In the following we assume that the state machine is executed by an interpreter that process an incoming stream of events, and traversing run-time pointers representing transitions accordingly, executing on-entry actions, where the notion of the current state and the values of the global variables are the collective run-time state of the program. The structure that the interpreter operates on can be described as derived  schema from the abstract syntax scheme shown on the right of Figure~\ref{LST:statemachines}, but specialized for a particular machine, e.g., \lstinline{Doors}.
This schema will have additional fields and associations modeling the run-time data required for execution. For the \lstinline{Doors} statemachine, this means adding the current state to the \lstinline{Machine} type (required for any machine) and the machine specific data fields, namely \lstinline{isClosed}. The run-time schema for \lstinline{Doors} is then the following\footnote{It is possible to represent the global variables at run-time using an untyped table or hashmap, but for the sake of schema migration complexity, the encoding as fields in classes is more instructive.}:
\begin{quote}
\begin{lstlisting}[language=java,morekeywords={on},mathescape=true]
class Machine$^{++}$
  states: State*
  current: State
  isClosed: bool
\end{lstlisting}
\end{quote}
This \textit{extended} class $\text{\lstinline{Machine}}^{++}$ also represents state machines, but this time at run time.

@@

\section{Live programming in the context of Elm architecture}
\label{SECT:elm}
\subsection{Context}
Another challenge involving stateful live programming can be drawn from the programming model based on the Elm architecture. In this model, a reactive (web) application is structured in terms of current state and events that affect the state. The implementation then consists of two functions that we may call \texttt{update} and \texttt{render}:

\begin{quote}
\begin{lstlisting}[language=ml,morekeywords={on}]
type State = { .. }
type Event = .. | ..

val update : State -> Event -> State
val render : State -> Html
\end{lstlisting}
\end{quote}

The programming model works as follows:
\begin{itemize}
    \item \texttt{State} represents the entire application state, i.e., everything that the user can work with.
    \item \texttt{Event} represents all events that the user can trigger by interacting with the application.
    \item \texttt{update} is called whenever an event happens. It takes the current state, the event and computes a new state.
    \item \texttt{render} takes the current state and produces a representation of what should be displayed on the screen (e.g., an HTML tree).
\end{itemize}

Traditionally, when Elm applications are developed, they are restarted each time the code is modified and any previous state is discarded. However, a more effective programming model based on live programming would allow live updates to both the code of the two functions and the structure of the two types.

As an example, consider the uninspiring, but well-known, TODO list application. The types that capture the state and events in the application may look as follows:

\begin{quote}
\begin{lstlisting}[language=ml,morekeywords={on}]
type Item = { id : id; title : string; completed : bool }
type State = { items : Item list }
type Event =
  | SetCompleted of id * bool
  | SetTitle of id * string
  | Remove of id
  | Add of string
\end{lstlisting}
\end{quote}

The application state consists of a list of items. Each item has a unique ID alongside with a title and a flag indicating whether it is completed. The events represent edits to the items, deletion and addition. The implementation of the \texttt{update} and \texttt{render} functions is simple and not important for the challenge.

\subsection{Challenge}
Now, imagine that we have a running TODO list application with the above state and events. To test the application, the programmer has already created a number of items and so there is a single value of the \texttt{State} type that represents a current state of the application such as:

\begin{quote}
\begin{lstlisting}[language=ml,morekeywords={on}]
{ items : [
   [ { id = 1; title = "check twitter"; completed = true  }
     { id = 1; title = "Write paper"; completed = false } ] }
\end{lstlisting}
\end{quote}

As above, there are a number of edits to the code and types that the programmer may want to do without restarting the application. Those are summarized in Table~\ref{TBL:elmchanges}. Modifying the code is simple and only requires waiting until the current execution completes. Modifying the \texttt{Event} type is also simple, but it may lead to unused code or missing case in \texttt{update} that needs to be addressed. Finally, modifying \texttt{State} ranges from relatively simple problems (adding a new field) to challenging case when the structure of \texttt{State} is changed.

For example, imagine that the programmer would want to edit the structure of the state and migrate the previous definition of \texttt{State} to the following new definition where individual fields are stored in separate lists:

\begin{quote}
\begin{lstlisting}[language=ml,morekeywords={on}]
type State =
  { ids : id list
    titles string list
    completes : bool list }
\end{lstlisting}
\end{quote}

This representation is semantically equivalent (assuming the lists have the same length) to the original one. It should thus, in principle, be possible to migrate the original state value to a value using the new structure. Moreover, it should, in principle, be also possible to automatically transform the implementations of \texttt{update} and \texttt{render} to work as before, but using the new state structure.

\begin{table}[t]
  \centering
\begin{tabular}{lll}\toprule
\textsc{Change} & \textsc{What} & \textsc{How}\\\midrule
modify & \texttt{render} & simple, but wait until current execution finishes  \\
modify & \texttt{update} & simple, but wait until current execution finishes  \\
add & case to \texttt{Event} & requires adding corresponding case to \texttt{update} \\
remove & case from \texttt{Event} & remove unused code from \texttt{update} \\
add & field to \texttt{State} & migrate state value and initialize field \\
remove & field from \texttt{State} & migrate state (assuming field unused) \\
modify & structure of \texttt{State} & migrate state value and edit code accordingly \\
\bottomrule
\end{tabular}
\caption{Possible program changes and how to deal with them at run time}
\label{TBL:elmchanges}
\end{table}

@@

\begin{quote}
\begin{lstlisting}[language=ml]
type Item = { id : id; title : string; assignee : string }
\end{lstlisting}
\end{quote}

@@

\begin{quote}
\begin{lstlisting}[language=ml]
type Item = { id : id; title : string; assignees : Array<string> }
\end{lstlisting}
\end{quote}

@@

\begin{quote}
\begin{lstlisting}[language=ml]
item.assignees = if (item.assignee == null) then [] else [item.assignee]
\end{lstlisting}
\end{quote}

@@

\begin{quote}
\begin{lstlisting}[language=ml]
todoScalar = { id: 1, title: "Foo", assignee: A }
todoArray = { id: 1, title: "Foo", assignees: [A, B] }

todoScalar.assignee = C
\end{lstlisting}
\end{quote}

>>>>

\usepackage{csquotes}
\usepackage{changepage}
\usepackage{multicol}
\usepackage{ccicons}
\usepackage[many]{tcolorbox}
\usepackage{tikz-cd}

\lstdefinelanguage[programming]{TeX}[AlLaTeX]{TeX}{%
  deletetexcs={title,author,bibliography},%
  deletekeywords={tabular},
  morekeywords={abstract},%
  moretexcs={chapter},%
  moretexcs=[2]{title,author,subtitle,keywords,maketitle,titlerunning,authorinfo,affiliation,authorrunning,paperdetails,acks,email},
  moretexcs=[3]{addbibresource,printbibliography,bibliography},%
}%
\lstset{%
  language={[programming]TeX},%
  keywordstyle=\firamedium,
  belowskip=1em,
  aboveskip=1em,
  framesep=0.5em,
  basicstyle=\normalsize\ttfamily,
  %stringstyle=\color{RosyBrown},%
  %texcsstyle=*{\color{Purple}\mdseries},%
  %texcsstyle=*[2]{\color{Blue1}},%
  %texcsstyle=*[3]{\color{ForestGreen}},%
  %commentstyle={\color{FireBrick}},%
  numbersep=1em,
  xleftmargin=1.5em,
  escapechar=`,}
\newcommand*{\CTAN}[1]{\href{http://ctan.org/tex-archive/#1}{\nolinkurl{CTAN:#1}}}
%%

\newcounter{challengen}
\newcommand{\challenge}[1]{\subsection*{\addtocounter{challengen}{1}Challenge \#\thechallengen: #1}}

\DeclareRobustCommand{\frameworkbox}[2][gray!15]{
\begin{tcolorbox}[breakable,left=3pt,right=3pt,top=3pt,bottom=3pt,colback=#1,colframe=#1,parbox=false,
  width=\dimexpr\textwidth\relax,enlarge left by=0mm,boxsep=5pt,arc=0pt,enlarge top by=0.5em,%enlarge bottom by=0.0em,
  outer arc=0pt]\setlength{\parskip}{0.5em}\setlength{\parindent}{0em}{\firamedium Framework perspective.}\quad #2
\end{tcolorbox}}

@@

Many past improvements in the practice of programming worked by speeding up feedback loops,
both in collaboration with other programmers (through collaborative development tools
\cite{goldman2011real,kurniawan2015coder,replit}) and in individual's interaction with the
programming environment (through live programming \cite{rein2018exploratory}).

To make programming more effective, future programming systems will need to further tighten
the feedback loop. They will need to enable programmers to collaborate at a fine-grained level,
letting them adopt individual changes created by their collaborators (as in local-first software
\cite{localfirst}). To provide immediate feedback, future programming systems will also need to
transcend treating state as ephemeral, tied to the lifetime of a running program and reset when the
program is edited.

To conceive of such systems, we need to shift our focus to \emph{programming systems} \cite{techdims}
where program state is persistent and co-evolves with the behavior and data of the system. This is
already the case in collaborative applications with state persistent in a database, image-based
systems like Smalltalk and live programming environments with hot-reloading where the program can
be edited without resetting all state.

While we are firmly convinced that more collaborative and live programming systems are the
future of programming, there are a number of problems that need to be resolved in order for
such systems to become a reality. The key problem is \emph{type evolution}, i.e., how to
handle the change of schema of the data program works with in a way that preserves integrity
constraints, existing data and the logic of code.

Type evolution is not a new problem. It exists in various forms in a range of existing systems.
The problem of schema evolution in databases is well-studied \cite{erhard06}.
It typically considers data integrity, although co-evolution of database schema
and code has also been studied \cite{Cleve2006,wang2019synthesizing}.
In live programming, the problem is known as hot code reloading \cite{barenz2020essence,beckmann2021shortening},
although it typically focuses on updating transient state and ignores the problem of more complex
data migrations. The problem is also studied in the bidirectional transformation (BX)
community \cite{czarnecki2009bidirectional}, in work on model-driven development
\cite{metamodels,alanen2003difference} and in more low-level work on dynamic software updating
\cite{hicks2005dynamic}.

Looking at type evolution from the perspective of programming systems can offer a new perspective that
unfies the many different aspects of the problem partially explored in other areas. A programming
system may have permanent data (as in a database) whose structure can change, requiring
corresponding change to data and code. A programming system may have a transient state conforming
to a type structure that may change during live programming. A programming system may also need
to allow programmers to independently work on different versions of a program.

In this paper, we draw connections between these different kinds of type evolution, pointing out
an essential shared structure. First, we present a shared conceptual
framework for thinking about the relationship between schema, code, and data in a programming
system. Second, we present five case studies that use the framework to study type evolution in a
highly diverse set of real-world programming scenarios. We hope this will encourage knowledge
sharing across different subfields, motivate further research on this important problem and
offer inspiring challenges that developers of future live and collaborative programming systems
will attempt to tackle.

\begin{figure}[t]
\centering
\vspace{-1em}
\includegraphics[width=0.45\textwidth]{figures/layers.png}
\caption{Pace layers from Stewart Brand's \emph{How Buildings Learn} \cite{Brand95}.
  Different layers of a building change at different pace. The geographic
  site of a building is eternal, its structure remains stable for decades, space plan
  changes every couple of years.}
\label{fig:layers}
\end{figure}

\section{Type Evolution}
To paraphrase Stewart Brand \cite{Brand95}, ``almost no programs evolve well. But all programs
(except for monuments) evolve anyway, however poorly, because the usages in and around them are
changing constantly.'' Programs evolve both while they are being created and during their operation
in response to new needs. As with buildings, different aspects of program evolve at different pace
(Figure~\ref{fig:layers}). And ``because of the different rates of change of its components,
a program is always tearing itself apart.''

\subsection{Types, Data and Code}

As is the case for buildings, making a change at surface layers (skin or stuff) is relatively easy,
but a change at a more fundamental layer (structure) has cascading effects on layers that depend on it.
In the case of programs, a more fundamental layer defines the shape of data and code.
We refer to this structure as \emph{types}, although we do not use the term in an precise technical
sense. In our conceptual framework, types may be data types that define the shape of data, they
may be explicitly declared in code and checked statically, as well as dynamically checked or
even implicit. The key property of types for us is that they determine the shape of other layers
of a program, namely data and code. As a result, a change of program types requires a
corresponding change in program data and code.

@@

The second dimension of challenges that we consider is non-local. A program may have multiple
variants that exist independently and need to be synchronized. The difficulty of the problem
depends on what kind of synchronization we want to support. If all variants of the program should
eventually adopt all changes (converge), the complexity is lesser. If we allow user to apply
only certain changes from other users (divergence), the complexity is greater. A typical case
may be a combination of the two where types and schema of a program should converge, but data
may diverge.

\begin{figure}[t]
\centering
\includegraphics[width=0.6\textwidth]{figures/arrows.png}
\caption{When a type changes from $T_1$ to $T_2$, a corresponding change needs to take place
  in data and code whose shape is defined by type. The challenge is finding a suitable
  corresponding transformation (semi-)automatically.}
\label{fig:change}
\end{figure}

@@

We now describe a conceptual framework lets us view a range of diverse set of real-world
type evolution problems in a unified way. We distinguish between two dimensions.
The \emph{local dimension} is concerned with how type evolution affects data and code within
a single program. The \emph{non-local dimension} is concerned with how type evolution
propagates across different program variants that may exist independently.

\paragraph{Program Layers.}
We consider the more permanent layer of \emph{types} and two less permanent layers of code and
data that are structured according to types. In specific case studies, the layers may map to
different aspects of programs, some may be not present and some may be implicit.

\begin{itemize}
\item \emph{Types} -- structure program data and determine aspects of code. A change in types
  requires a corresponding change in data and code. Types may be explicit (database schema, type definition)
  or implicit (shape of data structures, shape of a document).
\item \emph{Data} -- information stored by the program. Data may be more permanent (data in a
  database) or less permanent (state in live programming). When types evolve, data needs to be
  updated or (in case of transient state) discarded.
\item \emph{Code} -- or program logic, but not including source code that defines types.
  When types evolve, code needs to change to consume/produce data of a correct shape.
\end{itemize}

\paragraph{Program Variants.}
In programming systems that enable collaboration, multiple variants of a program may exist
concurrently and their types, data and code may need to be synchronized. Although collaboration
is missing from some of our case studies, it adds an important dimension that is crucial for
type evolution in future programming systems. We consider two characteristics.

\begin{itemize}
\item \emph{Convergence} vs. \emph{Divergence}. In the convergence model, all program variants
  eventually adopt all changes. It may not be possible to adopt a change without adopting
  all earlier changes. In the divergence model, a user may choose only
  particular changes they want to adopt, but keep other aspects of their custom design.
\item \emph{Centralized} vs. \emph{Decentralized}. In the centralized model, changes (type
  evolution) can only originate from a particular source. In the decentralized model,
  changes can be done on any of the multiple co-existing variants of a program.
\end{itemize}

\paragraph{Challenge Problems.}
The conceptual framework offers a wide range of configurations for challenge problems, but
our aim is not to enumerate all options. We use it to recast a number of
existing challenges in real-world programming systems as instances of the same problem of
type evolution. The case studies show how well-known solutions or problems in one domain
map to another domain and they also provide and inspiration and a benchmark for
developers of future programming systems.

\section{Live programming in the Elm architecture}
\label{SECT:elm}

The first challenge problem we recast in terms of our conceptual framework is live programming
of user interfaces based on the Elm architecture. In this model, a reactive web application is
structured in terms of current state and events that affect the state. The implementation then
consists of two functions called \texttt{update} and \texttt{render}:

\begin{lstlisting}[language=ml,morekeywords={on}]
type State = { .. }
type Event = .. | ..

val update : State -> Event -> State
val render : State -> Html
\end{lstlisting}

\noindent
The programming model works as follows:

\begin{itemize}
\setlength\itemsep{0em}
\item \texttt{State} represents the application state, i.e., everything that the user can work with.
\item \texttt{Event} represents events that the user can trigger by interacting with the application.
\item \texttt{update} is called when an event happens and computes a new program state.
\item \texttt{render} takes the current state and produces a visual representation of the page.
\end{itemize}

\noindent
As an example, consider the well-known TODO list app. Its state consists of a list of items,
each with a unique ID, a title and a flag indicating whether it has been completed.
The events represent changing of an item, deletion and addition:

\begin{lstlisting}[language=ml]
type Item = { id : id; title : string, completed : bool }
type State = { items : Item list }
type Event =
  | SetTitle of id * string
  | SetCompleted of id * bool
  | Remove of id
  | Add of string
\end{lstlisting}

\noindent
Some programming systems that use the Elm architecture support hot reloading when the
implementation of the \texttt{render} or \texttt{update} function changes, but they typically
discard state and restart the application when types are changed. A more effective programming
system based on live programming would allow updates to both the code of the two functions and
the structure of the two types.

\begin{table}[t]
\begin{tabular}{lll}\toprule
{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
add & case to \texttt{Event} & requires adding corresponding case to \texttt{update} \\
remove & case from \texttt{Event} & remove unused code from \texttt{update} \\
add & field to \texttt{State} & migrate state value and initialize field \\
remove & field from \texttt{State} & migrate state (assuming field unused) \\
modify & structure of \texttt{State} & migrate state value and edit code accordingly \\
\bottomrule
\end{tabular}
\vspace{0.3em}
\caption{Changes to types and how a programming system should handle them}
\label{TBL:elmchanges}
\vspace{-1em}
\end{table}

\challenge{Live Type Evolution of State and Events}
Assume that we have a running TODO list with the above state and events. To test the application,
the programmer has already created a number of items and so there is a value of the
\texttt{State} type that represents the current state of the application such as:

\begin{lstlisting}[language=ml,morekeywords={true,false}]
{ items : [
   { id = 1; title = "Check Twitter"; completed = true }
   { id = 2; title = "Write the paper"; completed = false } ] }
\end{lstlisting}

\noindent
There is a number of edits to the types that the programmer may now want to do without
restarting the application and losing the data. For example, they may want to change the
\texttt{completed} field of \texttt{Item} to instead store optional completion time:

\begin{lstlisting}[language=ml]
type Item = { id : id; title : string; completed : DateTime option }
type State = { items : Item list }
type Event =
  | .. | SetCompleted of id * DateTime option | ..
\end{lstlisting}

\noindent
To apply the change without restarting the application, the existing data need to be migrated
to match the new type definition. In this case, this likely cannot be done fully automatically
and the programmer may need to provide mapping (\texttt{false} to \texttt{None}, \texttt{true}
to \texttt{Some(DateTime(2024,5,1))}). If the programmer changes \texttt{Item} and \texttt{SetCompleted}
at the same time, the \texttt{update} function likely does not need to change, but the rendering
code needs to be modified (the programmer needs to specify how to map \texttt{DateTime option}
back to a Boolean for a checkbox).

More generally, there is a number of edits to the types that the programmer may want to do
without restarting the application. The edits and the desired handling in the programming
system are summarized in Table~\ref{TBL:elmchanges}. Modifying the \texttt{Event} type does not
require data migration (assuming that the change does not happen while there are unprocessed
events). It may result in unused code or missing cases in \texttt{update} that need to be addressed.
Modifying \texttt{State} ranges from relatively simple problems (e.g., adding a new field with a
default value) to challenging case. In particular, if a programer refactors \texttt{State} to
semantically equivalent type, the programming system should, in principle, be able to automatically
migrate both the original value and implementation of both functions to match the new type.

@@

@@

\textcolor{red}{TODO: The rest needs more editing}

\frameworkbox{
In this
}

\newpage
\textcolor{red}{TODO: The rest needs more editing}

@@

\section{diagrams}

  \begin{tikzcd}[sep = huge]
    \textsf{code}_1
    \arrow[d, "\textsf{refactor}"{left},""{name=L}]
    &
    \textsf{schema}_1
    \arrow[l, "\textsf{constrains}"{above}]
    \arrow[r, "\textsf{constrains}"]
    \arrow[d, red, "\textsf{change}"{near start, description}, ""{name=CL, left}, ""{name=CR}]
    &
    \textsf{data}_1
    \arrow[d, ""{name=R, left}, "\textsf{migrate}"{right}]
    \\
    \textsf{code}_2
    &
    \textsf{schema}_2
    \arrow[l, "\textsf{constrains}"{below}]
    \arrow[r, "\textsf{constrains}"{below}] &
    \textsf{data}_2
    \arrow[Rightarrow, from=CL, to=L, "\textsf{propagate}"{below}]
    \arrow[Rightarrow, from=CR, to=R, "\textsf{propagate}"{below}]
  \end{tikzcd}

@@

In the following we assume that the state machine is executed by an interpreter that process an incoming stream of events, and traversing run-time pointers representing transitions accordingly, executing on-entry actions, where the notion of the current state and the values of the global variables are the collective run-time state of the program. The structure that the interpreter operates on can be described as derived  schema from the abstract syntax scheme shown on the right of Figure~\ref{LST:statemachines}, but specialized for a particular machine, e.g., \lstinline{Doors}.
This schema will have additional fields and associations modeling the run-time data required for execution. For the \lstinline{Doors} statemachine, this means adding the current state to the \lstinline{Machine} type (required for any machine) and the machine specific data fields, namely \lstinline{isClosed}. The run-time schema for \lstinline{Doors} is then the following\footnote{It is possible to represent the global variables at run-time using an untyped table or hashmap, but for the sake of schema migration complexity, the encoding as fields in classes is more instructive.}:

\begin{lstlisting}[language=java,morekeywords={on},mathescape=true]
class Machine$^{++}$
  states: State*
  current: State
  isClosed: bool
\end{lstlisting}

This \textit{extended} class $\text{\lstinline{Machine}}^{++}$ also represents state machines, but this time at run time.

@@

@@

\begin{lstlisting}[language=ml]
type Item = { id : id; title : string; assignee : string }
\end{lstlisting}

@@

\begin{lstlisting}[language=ml]
type Item = { id : id; title : string; assignees : Array<string> }
\end{lstlisting}

@@

\begin{lstlisting}[language=ml]
item.assignees = if (item.assignee == null) then [] else [item.assignee]
\end{lstlisting}

@@

\begin{lstlisting}[language=ml]
todoScalar = { id: 1, title: "Foo", assignee: A }
todoArray = { id: 1, title: "Foo", assignees: [A, B] }

todoScalar.assignee = C
\end{lstlisting}
